Low Complexity Adaptive Beamformer\\ for Active Sonar Imaging

### 2015-08-29 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:08:29	19:52:25	(Sat29.Aug)

MVDR seeks to minimize the power of the beamformer, under the constraint of a unity gain in the desired direction $\phi$~\cite{Capon1969}:
%
\begin{align}
\underset{\w[n]}{\argmin}\, E\{|z[n]|^2\} &= \underset{\w[n]}{\argmin}\, \w[n]\R[n]\w\H[n]\nn
\text{subject to } \w[n]\a_\phi &= 1.\label{eq_mvdr_definition}
\end{align}
%
Here $\a_\phi$ is a steering vector, $E\{\cdot\}$ is the expectation operator, and $\R[n] = E\{\x\x\H\}$ is the spatial covariance matrix. This is a convex optimization problem with the solution:
%
\begin{gather}
\vec w[n] = \frac{\Ri[n]\a_\phi}{\a_\phi\T\Ri[n]\a_\phi},\label{weights}
\end{gather}
where $\R=E\{\x[n]\x\H[n]\}\in\mathbb{C}^{M,M}$ is the spatial covariance matrix for the full array. The problem lies in estimating and inverting this spatial covariance matrix. Spatial averaging is needed to avoid signal cancellation~\cite{Kailath1985}, temporal averaging is needed to obtain true speckle statistics~\cite{Synnevag2009a}, and diagonal loading is needed to improve robustness to parameter errors~\cite{Cox1987,Maksym1979}. The steps are also needed to ensure that the covariance matrix is numerically well conditioned and hence invertible.

@@

A much less complex alternative to MVDR is the low complexity adaptive (LCA) beamformer. It iterates through a set of $P$ windows and selects the window $p$ that best fulfills the minimum variance criteron:
%
\begin{align}
\underset{p}{\argmin}\, E\{|z[n]|^2\} = \underset{p}{\argmin}\, E\big\{|\w_{p}\H\x[n]|^2\big\}\nn
\qquad\text{subject to}\qquad \w\H\a_\phi = 1.\label{eq_lca_definition}
\end{align}\label{lca_criterion}
%
Note how closely LCA is related to the MVDR definition in \eq{eq_mvdr_definition}. The optimization criterion and constraint is the same, but LCA has a finite and discrete solutionspace for the weights. As will be demonstrated in section \ref{lca_window_function}, this can be done because a robustified MVDR implementation seems to be constrained to window functions similar to the conventional ones.

We can estimate $E\big\{|\w_p\H\x[n]|^2\big\}$ as
%
\begin{align*}
\sigma^2_{z_p}[n] = \frac{1}{N_k} \sumb{n'=n-K}{n+K} | z_p[n'] |^2
\end{align*}
%
which is the sample covariance averaged over $N_k=2K+1$ temporal samples. To add some extra flexibility, we can also estimate the variance by a weighted combination of local pixels
%
\begin{align*}
\sigma^2_{z_p}[x,n] = \frac{1}{N_x N_k} \sumb{x'=x-X}{x+X} \sumb{n=n-K}{n+K} w_\text{prf}[x',n']\big| z_p[x',n'] \big|^2
\end{align*}
%
where $w_\text{prf}[x',n']$ is a normalized 2 dimensional weight function.

@@

As will be demonstrated in upcoming sections, the LCA beamformer works very well with windows generated from the Kaiser-Bessel function. Let it be expressed on vector form as:
%
\begin{align*}
\f_\beta = \bmat{
f_0(\beta) \\
\vdots\\
f_{M-1}(\beta)
}
\end{align*}
%
where
%
\begin{align}
f_m(\beta) = \frac{I_0\left(\pi\beta\sqrt{1-\left(\frac{2m}{M-1}-1\right)^2}\right)}{I_0(\pi\beta)}
\end{align}
%
and $I_0$ is the zeroth order modified Bessel function of the first kind:
%
\begin{align}
I_0(x) = \sumb{a=0}{\infty} \left[ \frac{\left(\frac{x}{2}\right)^a}{a!} \right]^2.
\end{align}
%
The Kaiser-Bessel window is near optimal in the sense of having its peak energy concentration around $\theta=0^\circ$, for a given time-bandwidth product specified with the parameter $\beta$:
%
\begin{align}
\beta = \frac{TB}{2},
\end{align}
%
where $T$ is the extent of the window in time and $B$ is its bandwidth. Adjusting $\beta$ changes the trade-off between mainlobe width and sidelobe level. When $\beta=0$ the window becomes a rectangular, while at large values ($\beta>5$) the window converges to a Gaussian both in time and frequency. This class of windows is generally considered well suited for separating closely spaced sources with amplitudes of a high dynamic range~\cite{Harris1978}, they are easy to make, and they are optimal for any value of $\beta$.

@@

To allow LCA some added flexibility we also create variations of each window that are steered a little to each side.
%
\begin{align}
\w_{\beta,\phi} = \frac{\f_\beta\H\diag{(\a_\phi)}}{\f_\beta\H\a_\phi}\\
\end{align}
%
where
%
\begin{align}
\a_\phi = \bmat{
1 \\
e^{-j\frac{md}{\lambda}\sin(\phi)} \\
\vdots\\
e^{-j\frac{(M-1)d}{\lambda}\sin(\phi)}
}.
\end{align}
%
Here $d$ is the element spacing, $\lambda$ is the wavelength and $\phi$ is the steering amount. The normalization factor $\f_\beta\H\a_\phi$ is the reciprocal of the window's coherent gain. It ensures unit gain in the direction of interest as required by (\ref{eq_lca_constraint}). Since the signal-to-noise ratio is constant for a specific window (same $\beta$-value), this normalization also proportionally increases the incoherent noise gain.

@@

There is an upper bound to the amount of steering we should apply to a window. Since each window's response is constrained to a unit gain in the look direction, each steered window's response is gained as shown in \Fig{windows_steering}. This increases the white noise gain and deteriorates the signal-to-noise ratio. To limit this we placing an upper bound to the steering, taking into account that a wide window can be allowed more steering than a narrow one.

>>>>

MVDR seeks to minimize the power of the beamformer, under the constraint of a unity gain in the desired direction $\phi$~\cite{Capon1969}:
%
\begin{align}
\underset{\w[n]}{\argmin}\, E\{|z[n]|^2\} &= \underset{\w[n]}{\argmin}\, \w[n]\R[n]\w\H[n]\nn
\text{subject to } \w[n]\a_\phi &= 1.\label{eq_mvdr_definition}
\end{align}
%
Here $\a_\phi$ is a steering vector, $E\{\cdot\}$ is the expectation operator, and $\R[n] = E\{\x\x\H\}$ is the spatial covariance matrix. This is a convex optimization problem with the solution:
%
\begin{gather}
\vec w[n] = \frac{\Ri[n]\a_\phi}{\a_\phi\T\Ri[n]\a_\phi},\label{weights}
\end{gather}
where $\R=E\{\x[n]\x\H[n]\}\in\mathbb{C}^{M,M}$ is the spatial covariance matrix for the full array. The problem lies in estimating and inverting this spatial covariance matrix. We have described the exact steps in~\cite{Buskenes2014}. To avoid signal cancellation we apply spatial averaging by computing a mean covariance matrix from a set of subarrays with length $L$~\cite{Kailath1985}. For true speckle statistics we perform temporal averaging over $N_k = 2K+1$ temporal samples~\cite{Synnevag2009a}. Finally, to improve robustness to parameter errors we apply $d$ percent diagonal loading~\cite{Cox1987,Maksym1979}. The steps are also needed to ensure that the covariance matrix is numerically well conditioned and hence invertible.

@@

A much less complex alternative to MVDR is the low complexity adaptive (LCA) beamformer. It iterates through a set of $P$ windows and selects the window $p$ that best fulfills the minimum variance criteron:
%
\begin{align}
\underset{p}{\argmin}\, E\{|z[n]|^2\} = \underset{p}{\argmin}\, E\big\{|\w_{p}\H\x[n]|^2\big\}\nn
\qquad\text{subject to}\qquad \w\H\a_\phi = 1.\label{eq_lca_definition}
\end{align}\label{lca_criterion}
%
Note how closely LCA is related to the MVDR definition in \eq{eq_mvdr_definition}. The optimization criterion and constraint is the same, but LCA has a finite and discrete solutionspace for the weights. As will be demonstrated in section \ref{lca_window_function}, LCA will perform similarly to MVDR because a robustified MVDR implementation seems to be constrained to window functions similar to the ones we let LCA choose from.

In practice, we estimate the beamformer output power as
%
\begin{align*}
E\big\{|z[n]|^2\big\} = \frac{1}{N_k} \sumb{n'=n-K}{n+K} | z_p[n'] |^2
\end{align*}
%
which computes the sample power average over $N_k=2K+1$ temporal samples. This is the same as we did for MVDR. Our signal is only a few samples long after matched filtering so we have set $K=1$ for both MVDR and LCA.

If extra flexibility is desired, we can also estimate the variance by a weighted combination of local pixels
%
\begin{align*}
\sigma^2_{z_p}[x,n] = \frac{1}{N_x N_k} \sumb{x'=x-X}{x+X} \sumb{n=n-K}{n+K} w_\text{prf}[x',n']\big| z_p[x',n'] \big|^2
\end{align*}
%
where $N_x = 2X+1$ is the number of azimuth lines to average over, and $w_\text{prf}[x',n']$ is a normalized 2 dimensional weight function.

@@

As will be demonstrated in upcoming sections, the LCA beamformer works very well with windows generated from the Kaiser-Bessel function. We will express it in vector form as:
%
\begin{align*}
\f_\beta = \bmat{
f_0(\beta) \\
\vdots\\
f_{M-1}(\beta)
}
\end{align*}
%
where
%
\begin{align}
f_m(\beta) = \frac{I_0\left(\pi\beta\sqrt{1-\left(\frac{2m}{M-1}-1\right)^2}\right)}{I_0(\pi\beta)}
\end{align}
%
and $I_0$ is the zeroth order modified Bessel function of the first kind:
%
\begin{align}
I_0(x) = \sumb{a=0}{\infty} \left[ \frac{\left(\frac{x}{2}\right)^a}{a!} \right]^2.
\end{align}
%
The Kaiser-Bessel window is near optimal in the sense of having its peak energy concentration around $\theta=0^\circ$, for a given time-bandwidth product specified with the parameter $\beta$:
%
\begin{align}
\beta = \frac{TB}{2},
\end{align}
%
where $T$ is the extent of the window in time and $B$ is its bandwidth. Adjusting $\beta$ changes the trade-off between mainlobe width and sidelobe level. When $\beta=0$ the window becomes a rectangular, while at large values ($\beta>5$) the window converges to a Gaussian both in time and frequency. This class of windows is generally considered well suited for separating closely spaced sources with amplitudes of a high dynamic range~\cite{Harris1978}, they are easy to make, and they are optimal for any value of $\beta$.

@@

To allow LCA some added flexibility we also create variations of each window that are steered a little to each side.
%
\begin{align}
\w_{\beta,\phi} = \frac{\f_\beta\H\diag{(\a_\phi)}}{\f_\beta\H\a_\phi}\\
\end{align}
%
where
%
\begin{align}
\a_\phi = \bmat{
1 \\
e^{-j\frac{md}{\lambda}\sin(\phi)} \\
\vdots\\
e^{-j\frac{(M-1)d}{\lambda}\sin(\phi)}
}.
\end{align}
%
Here $d$ is the element spacing, $\lambda$ is the wavelength and $\phi$ is the steering amount. The normalization factor $\f_\beta\H\a_\phi$ is the reciprocal of the window's coherent gain. It ensures unit gain in the direction of interest as required by (\ref{eq_lca_definition}). Since the signal-to-noise ratio is constant for a specific window (same $\beta$-value), this normalization also proportionally increases the incoherent noise gain.

@@

There is an upper bound to the amount of steering we should apply to a window. Since each window's response is constrained to a unit gain in the look direction, the steered window will have increased white noise gain and deteriorated signal-to-noise ratio as shown in \Fig{windows_steering}. To limit this we placing an upper bound to the steering, taking into account that a wide window can be allowed more steering than a narrow one.
